Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-05 slides/slides-10/notes-10.tex
more changes

\begin{definition}[Yoneda Embedding]
  Let $\cat A$ be locally small. The \emph{Yoneda embedding} of $\cat A$ is the functor
  \begin{align*}
    H_-\from \cat A\to[\cat A^\op,\Set]
  \end{align*}
  defined by
  \begin{itemize}
  \item $H_-(A)=H_A$ for objects $A\in\cat A$ and
  \item for maps $A\xrightarrow{f}A'$, $H_-(f)$ is the natural transformation defined for $B\in \cat A$ by
    \begin{align*}
      H_A(B)=\cat A(B,A)&\to H_{A'}(B)=\cat A(B,A')\\
      g&\mapsto  f\of g **\\
    \end{align*}
  \end{itemize}
  \vspace{-0.8cm}
  \includegraphics{nattrans.pdf}
\end{definition}
Basically, for preorders ** constructs a proof from $p\leq p'$ that for all $q\leq p$ also $q\leq p'$ holds.

@@

\section{Yoneda Lemma}
To do so we state the following technical lemma due to Yoneda:
\begin{lemma}[Yoneda]
  Let $\cat A$ be locally small. Then
  \begin{align*}
    [\cat A^\op,\Set](H_A,F)\iso F(A)
  \end{align*}
  naturally in $A\in\cat A$ and $F\in [\cat A^\op,\Set]$. That is for each $B\xrightarrow{f}A$ in $\cat A$ and for each natural transformation $\theta \from F\to F'$ the following commute
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{-\of H_f} & [\cat A^\op,\Set](H_B,F) \\
    F(A) \ar[r]^{F(f)} \ar[u]^{\iso}& F(B) \ar[u]_{\iso}
                         } 
  \end{align*}
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{\theta\of -} & [\cat A^\op,\Set](H_A,F') \\
    F(A) \ar[r]^{\theta_A} \ar[u]^{\iso} & F'(A) \ar[u]^{\iso}
                             } 
  \end{align*}
\end{lemma}
It is going to help us understand what the arrows $F_A\to F_{A'}$ are (in the presheaf category). More generally, it states that an element of $F(A)$ ``is the same'' as a natural transformation $H_A\to F$.
\begin{proof}
  In order to show the isomorphism in $\Set$, which is supposed to be natural in both $A$ and $F$, we can proceed by:
  \begin{enumerate}
  \item Define a function $(\widehat{-})\from[\cat A^\op,\Set](H_A,F)\to F(A)$.
  \item Define a function $(\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)$.
  \item Show that $(\widehat{\widetilde{-}})=1_{F(A)}$.
  \item Show that $(\widetilde{\widehat{-}})=1_{[\cat A^\op,\Set](H_A,F)}$.
  \item Show that $(\widehat{-})$ is natural in $A$.
  \item Show that $(\widehat{-})$ is natural in $F$.
  \end{enumerate}
  It turns out that the definitions of $(\widehat{-})$ and $(\widetilde{-})$ are straight-forward as there is only one canonical choice that can possibly be made:
  \begin{enumerate}
  \item Let $\alpha\from H_A\to F$ be arbitrary. In order to define $(\widehat{-})\from[\cat A^\op,\Set](H_A,F)\to F(A)$, we have to come up with an element $\widehat\alpha\in F(A)$. However, $\alpha$ gives us a function $\alpha_A\from H_A(A)\to F(A)$ in $\Set$. Hence, we only have to come up with an element of $H_A(A)=\cat A(A,A)$ to obtain an element of $F(A)$ and we can simply choose $1_A$. Thus, we define $\widehat\alpha=\alpha_A(1_A)$.
  \item Let $a\in F(A)$. This time, we have to define a natural transformation $\widetilde a\from H_A\to F$, i.e. for each $B\in\cat A$ we have to define a function $\widetilde a_B\from H_A(B)\to F(B)$ in $\Set$ and show that the naturality condition is satisfied.

@@

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    F_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & F_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in F_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(f)\of F(g))(a)&F \text{ functor}\\
    &=F(f)(F(g)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

Now, it is easy to show that $H_-$ is full and faithful by instantiating $F$ with a covariant representable functor.
\begin{theorem}
  The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
\end{theorem}
\begin{proof}
  We have 
  \begin{align*}
    \widetilde H_f=(H_f)_A(1_A)=f\of 1_A=f.
  \end{align*}
  Hence,
  \begin{align*}
    \cat A(A,A')&\to[\cat A^\op,\Set](H_A,H_A')\\
    f&\mapsto H_f
  \end{align*}
  and
  \begin{align*}
    (\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)
  \end{align*}
  are the same function, which are bijective due to the proof of the Yoneda lemma (instantiating $F$ with $H_{A'}$).
\end{proof}

By a previous exercise we know the following:
\begin{lemma}
  Let $F\from\cat A\to\cat B$ be a functor and $A\xrightarrow{f} A'$ in $\cat A$.

@@

We start by proving the following ``distributivity''-lemma:

\begin{proposition}
  \label{prop:dis}
  Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C).
  \end{align*}
\end{proposition}

>>>>

\begin{definition}[Yoneda Embedding]
  Let $\cat A$ be locally small. The \emph{Yoneda embedding} of $\cat A$ is the functor
  \begin{align*}
    H_-\from \cat A\to[\cat A^\op,\Set]
  \end{align*}
  defined by
  \begin{itemize}
  \item $H_-(A)=H_A$ for objects $A\in\cat A$ and
  \item for maps $A\xrightarrow{f}A'$, $H_f$ is the natural transformation defined for $B\in \cat A$ by
    \begin{align*}
      H_A(B)=\cat A(B,A)&\to H_{A'}(B)=\cat A(B,A')\\
      g&\mapsto  f\of g **\\
    \end{align*}
  \end{itemize}
  \vspace{-0.8cm}
  \includegraphics{nattrans.pdf}
\end{definition}
Basically, for preorders ** constructs a proof from $p\leq p'$ that for all $q\leq p$ also $q\leq p'$ holds.

@@

\section{Yoneda Lemma}
To do so we state the following technical lemma due to Yoneda:
\begin{lemma}[Yoneda]
  Let $\cat A$ be locally small. Then
  \begin{align*}
    [\cat A^\op,\Set](H_A,F)\iso F(A)
  \end{align*}
  naturally in $A\in\cat A$ and $F\in [\cat A^\op,\Set]$. That is for each $B\xrightarrow{f}A$ in $\cat A$ and for each natural transformation $\theta \from F\to F'$ the following commute
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{-\of H_f} & [\cat A^\op,\Set](H_B,F) \\
    F(A) \ar[r]^{F(f)} \ar[u]^{\iso}& F(B) \ar[u]_{\iso}
                         } 
  \end{align*}
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{\theta\of -} & [\cat A^\op,\Set](H_A,F') \\
    F(A) \ar[r]^{\theta_A} \ar[u]^{\iso} & F'(A) \ar[u]^{\iso}
                             } 
  \end{align*}
\end{lemma}
It is going to help us understand what the arrows $H_A\to H_{A'}$ are (in the presheaf category). More generally, it states that an element of $F(A)$ ``is the same'' as a natural transformation $H_A\to F$.
\begin{proof}
  In order to show the isomorphism in $\Set$, which is supposed to be natural in both $A$ and $F$, we can proceed by:
  \begin{enumerate}
  \item Define a function $(\widehat{-})\from[\cat A^\op,\Set](H_A,F)\to F(A)$.
  \item Define a function $(\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)$.
  \item Show that $(\widehat{\widetilde{-}})=1_{F(A)}$.
  \item Show that $(\widetilde{\widehat{-}})=1_{[\cat A^\op,\Set](H_A,F)}$.
  \item Show that $(\widehat{-})$ is natural in $A$.
  \item Show that $(\widehat{-})$ is natural in $F$.
  \end{enumerate}
  It turns out that the definitions of $(\widehat{-})$ and $(\widetilde{-})$ are straight-forward as there is only one canonical choice that can possibly be made:
  \begin{enumerate}
  \item Let $\alpha\from H_A\to F$ be arbitrary. In order to define $(\widehat{-})\from[\cat A^\op,\Set](H_A,F)\to F(A)$, we have to come up with an element $\widehat\alpha\in F(A)$. However, $\alpha$ gives us a function $\alpha_A\from H_A(A)\to F(A)$ in $\Set$. Hence, we only have to come up with an element of $H_A(A)=\cat A(A,A)$ to obtain an element of $F(A)$ and we can simply choose $1_A$. Thus, we define $\widehat\alpha=\alpha_A(1_A)$.
  \item Let $a\in F(A)$. This time, we have to define a natural transformation $\widetilde a\from H_A\to F$, i.e. for each $B\in\cat A$ we have to define a function $\widetilde a_B\from H_A(B)\to F(B)$ in $\Set$ and show that the naturality condition is satisfied.

@@

    We still have to check naturality, i.e. that for any $B'\xrightarrow{g}B$ the following commutes:
    \begin{align*}
    \xymatrix{
    H_A(B)=\cat A(B,A) \ar[r]^{H_A(g)=-\of g} \ar[d]^{\widetilde a_B} & H_A(B')=\cat A(B',A) \ar[d]_{\widetilde a_B'} \\
    F(B) \ar[r]^{F(g)} & F(B').
                         } 
  \end{align*}
  Since this square is in $\Set$, it suffices to check $(\widetilde a_B'\of H_A(g))(f)=(F(g)\of\widetilde a_B)(f)$ for every $f\in H_A(B)=\cat A(B,A)$. Indeed, this holds:
  \begin{align*}
    (\widetilde a_B'\of H_A(g))(f)&=(F(f\of g))(a)\\
    &=(F(f)\of F(g))(a)&F \text{ functor}\\
    &=F(f)(F(g)(a))\\
    &=(F(g)\of\widetilde a_B)(f).
  \end{align*}
\item Let $a\in F(A)$. Clearly, we get $\widehat{\widetilde a}=a$:
  \begin{align*}
    \widehat{\widetilde a}=\widetilde a_A(1_A)=F(1_A)(a)=1_{F(A)}(a)=a.
  \end{align*}
\item[(d)-(f)] The remaining proof obligations are quite straight-forward to show and are thus left as an exercise.
  \end{enumerate}
\end{proof}

Now, it is easy to show that $H_-$ is full and faithful by instantiating $F$ with a covariant representable functor.
\begin{theorem}
  The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
\end{theorem}
\begin{proof}
  We have 
  \begin{align*}
    \widehat H_f=(H_f)_A(1_A)=f\of 1_A=f.
  \end{align*}
  Hence,
  \begin{align*}
    \cat A(A,A')&\to[\cat A^\op,\Set](H_A,H_A')\\
    f&\mapsto H_f
  \end{align*}
  and
  \begin{align*}
    (\widetilde{-})\from F(A)\to[\cat A^\op,\Set](H_A,F)
  \end{align*}
  are the same function, which are bijective due to the proof of the Yoneda lemma (instantiating $F$ with $H_{A'}$).
\end{proof}

By a previous exercise we know the following:
\begin{lemma}
  Let $F\from\cat A\to\cat B$ be a full and faithful functor and $A\xrightarrow{f} A'$ in $\cat A$.

@@

We start by proving the following ``distributivity''-property:

\begin{proposition}
  \label{prop:dis}
  Let $\cat A$ be cartesian closed such that it has coproducts. Then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C).
  \end{align*}
\end{proposition}

### 2017-07-05 slides/slides-10/slides-10.tex
Yet more changes.
more changes

\newtheorem{proposition}{Proposition}

@@

\begin{frame}
  \frametitle{Yoneda Embedding}
  Let $\cat A$ be locally small. The \emph{Yoneda embedding} of $\cat A$ is the functor
  \begin{align*}
    H_-\from \cat A\to[\cat A^\op,\Set]
  \end{align*}
  defined by
  \begin{itemize}
  \item $H_-(A)=H_A$ for objects $A\in\cat A$ and
  \item for maps $A\xrightarrow{f}A'$, $H_-(f)$ is the natural transformation defined for $B\in \cat A$ by
    \begin{align*}
      H_A(B)=\cat A(B,A)&\to H_{A'}(B)=\cat A(B,A')\\
      g&\mapsto  f\of g\\
    \end{align*}
  \end{itemize}
  \vspace{-0.8cm}
  \includegraphics{nattrans.pdf}
\end{frame}

@@

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle}
  \begin{proposition}
    Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
    \begin{align*}
      (A\times B)+(A\times C)\iso A\times (B+C).
    \end{align*}
  \end{proposition}
  \pause
  \begin{lemma}
    For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps
 all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle (Cont.)}
  \begin{proposition}
    Let $F,F'\from \cat A\to\cat B$ and $G\from\cat B\to\cat A$ such that $F\dashv G$ and $F'\dashv G$. Then $F\iso F'$
  \end{proposition}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category}
  \begin{proposition}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{proposition}
  \pause
  \begin{lemma}
    For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
    \begin{align*}
      \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
    \end{align*}
    for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category (Cont.)}  
  \begin{lemma}
    Let $F\from \cat A\to\Set$ be a presheaf. Then
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F
    \end{align*}
    for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
  \end{lemma}
  \pause
  \begin{proposition}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{proposition}
\end{frame}
\end{document}

>>>>

@@

\begin{frame}
  \frametitle{Yoneda Embedding}
  Let $\cat A$ be locally small. The \emph{Yoneda embedding} of $\cat A$ is the functor
  \begin{align*}
    H_-\from \cat A\to[\cat A^\op,\Set]
  \end{align*}
  defined by
  \begin{itemize}
  \item $H_-(A)=H_A$ for objects $A\in\cat A$ and
  \item for maps $A\xrightarrow{f}A'$, $H_f$ is the natural transformation defined for $B\in \cat A$ by
    \begin{align*}
      H_A(B)=\cat A(B,A)&\to H_{A'}(B)=\cat A(B,A')\\
      g&\mapsto  f\of g\\
    \end{align*}
  \end{itemize}
  \vspace{-0.8cm}
  \includegraphics{nattrans.pdf}
\end{frame}

@@

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle}
  \begin{theorem}
    Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
    \begin{align*}
      (A\times B)+(A\times C)\iso A\times (B+C).
    \end{align*}
  \end{theorem}
  \pause
  \begin{lemma}
    For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps
 all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle (Cont.)}
  \begin{theorem}
    Let $F,F'\from \cat A\to\cat B$ and $G\from\cat B\to\cat A$ such that $F\dashv G$ and $F'\dashv G$. Then $F\iso F'$.
  \end{theorem}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category}
  \begin{theorem}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{theorem}
  \pause
  \begin{lemma}
    For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
    \begin{align*}
      \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
    \end{align*}
    for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category (Cont.)}  
  \begin{lemma}
    Let $F\from \cat A\to\Set$ be a presheaf. Then
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F
    \end{align*}
    for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
  \end{lemma}
  \pause
  \begin{theorem}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{theorem}
\end{frame}
\end{document}
